---
title: "4323 Final Project- College Basketball Dataset"
author: "Ryan Nguyen and Justin George"
date: "2023-04-25"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Background on our Dataset

This is a dataset that contains information about college basketball teams from 2013 to 2019, including the team's performance statistics as well as their postseason performance. The dataset contains 3455 observations with 24 variables, including 5 categorical var4iables and 19 numerical variables. 

My project as a whole aims to use supervised learning to predict the stage at which a given college basketball team will be eliminated from the NCAA March Madness Tournament. The response variable is the categorical variable POSTSEASON, which represents the stage which the team was eliminated or where their season ended. We plan to use K-nearest neighbor classifier, support vector machines, and other categorical supervised learning algorithms to develop the classification model.



## Ryan: Conducting Principal Component Analysis (PCA)

Since we are using a large dataset with all the stats on college basketball teams during one season, and the high dimensionality of our data, we want to make sure that our PCA does its best to determine which variables account for the most variance to find the ones that are most significant. This will help predict our responses most accurately.

Here are some key points to note:\

- PCA will come in use for other two models we are running\
- PCA will be tested with and without effect\
- Some models with complete dataset; some models with limited amount\

## Slide with R Output

```{r cbb dataset}
library(readr)
#our data set
data <- read_csv("C:/Users/Ryan/Downloads/DATASCIENCESTSTS/PROJECT/cbb.csv")
data
#select only the numerical variables
num_vars <- c("G", "W", "ADJOE", "ADJDE", "BARTHAG", "EFG_O", "EFG_D", "TOR", "TORD", "ORB", "DRB", "FTR", "FTRD", "2P_O", "2P_D", "3P_O", "3P_D", "ADJ_T", "WAB")
#Standardize the data
data_std <- scale(data[, num_vars])
```

## Slide with Continued R Output

```{r results}
#Perform PCA
pca <- prcomp(data_std, center = TRUE, scale = FALSE)
#Print Summary of PCA
summary(pca)
#Plot cumulative proportion of variance explained by each principal component
plot(cumsum(pca$sdev^2 / sum(pca$sdev^2)), xlab = "Number of principal components", ylab = "Cumulative proportion of variance explained")
#Choose number of principal components to keep
num_components <- 5
#Compute the new dataset with the selected number of principal components
data_pca <- predict(pca, newdata = data_std)[, 1:num_components]
data_pca
#print summary of new dataset
summary(data_pca)
```
## Ryan: Results of my PCA
Using PCA, we found the top three principal components that explain the majority of the variance in our dataset, accounting for about 63.8% of the total variance. These components are related to offensive and defensive efficiency, and rebounding performance, which are key factors in determining a team's success in college basketball. Based on our results, our analysis shows that these components are highly correlated with each other, which indicates that teams that excel in one area are likely to excel in the others as well. 

Overall, our results suggest that PCA is a powerful tool for identifying the most important factors that contribute to a team's success in college  basketball.
